---
title: "Unequal and deteriorating sustainability trends over the past two decades"
author: "Alven Yu"
date: '2025-07-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(broom)
library(raster)
library(stars)
library(rasterVis)
library(sf)
library(sp)
library(rnaturalearth)
```

    This is the main code for article "Unequal and deteriorating sustainability trends over the past two decades". We select the most representative indicators to measure the sustaintability trends over the past two decades. The indicators are:
    
    * PM2.5 – as an environmental health indicator
    * CO2 – as an emission indicator
    * Nightlight – as SES indicator
    * NVDI – as a green health indicator
    

## 1. The change in indicator values per decade

```{r}
trend_function_update <- function(indictor = "co2") {
        
        grid_data_data_df <- readRDS(paste0(
                        "./grid_data_2000_2019/grid_",
                        indictor,
                        "_data_df_update.rds"
                ))
       
        # the lm for trend
        e1 <- Sys.time()
        trend_data <-
                grid_data_data_df %>% nest(data = -c(x, y)) %>% mutate(model = map(data, ~ lm(value ~ year, data = .)),
                                                                       tidied = map(model, tidy)) %>% unnest(tidied) %>%
                filter(term %in% c("year")) %>%
                dplyr::select(x, y, estimate,p.value)
        
        saveRDS(trend_data,
                file = paste0("./Results/data/trend_", indictor, "update2.rds"))
        e2 <- Sys.time()
        
        e2 - e1
        
        
        # clean the extreme values
        trend_data_clean <- trend_data %>%
                filter((
                        estimate >= quantile(trend_data$estimate, 0.0001, na.rm = TRUE)
                ) & (
                        estimate <= quantile(trend_data$estimate, 0.9999, na.rm = TRUE)
                )) %>%
                mutate(estimate = estimate * 10) %>% # per decade
                dplyr::select(x, y, estimate)
        
        raster_trend_data_clean <-
                raster::rasterFromXYZ(xyz = trend_data_clean, crs = "+proj=longlat +datum=WGS84 +no_defs")
        
        raster::writeRaster(
                x = raster_trend_data_clean,
                filename = paste0("./Results/", indictor, "_trend_global", "_update.tif"),
                format = "GTiff",
                overwrite = TRUE
        )
}
```

```{r}
trend_function_update(indictor = "PM2_5")

trend_function_update(indictor = "co2")

trend_function_update(indictor = "NDVI")

trend_function_update(indictor = "nightlight")

```

## 2. combined with indictors

```{r}
# load indictor raster
PM2_5_trend <- raster::raster("./Results/PM2_5_trend_global_update.tif")
co2_trend <- raster::raster("./Results/co2_trend_global_update.tif")
nightlight_trend <- raster::raster("./Results/nightlight_trend_global_update.tif")
NDVI_trend <- raster::raster("./Results/NDVI_trend_global_update.tif")

# crop to the same extent
new_extent <- extent(-180, 180, -55.4, 74)
co2_trend_crop <- raster::crop(co2_trend, new_extent)
nightlight_trend_crop <- raster::crop(nightlight_trend, new_extent)
NDVI_trend_crop <- raster::crop(NDVI_trend, new_extent)

raster_stack_co2_nl_ndvi <- stack(co2_trend_crop, nightlight_trend_crop, NDVI_trend_crop)

# convert to the same extent for PM2.5
PM2_5_trend_crop <- resample(PM2_5_trend, raster_stack_co2_nl_ndvi,  method='bilinear')

raster_stack_all <- stack(raster_stack_co2_nl_ndvi, PM2_5_trend_crop)

rm(co2_trend_crop, nightlight_trend_crop, NDVI_trend_crop, PM2_5_trend_crop)
rm(NDVI_trend, nightlight_trend, co2_trend, PM2_5_trend,raster_stack_co2_nl_ndvi)

# save the stack raster
# convert the raster to stars
raster_stack_all <- st_as_stars(raster_stack_all)

# save the stars as tif
write_stars(raster_stack_all, "./Results/raster_stack_all_update.tif")

```

## 3. classification of the trend

```{r}
# load stack raster
raster_stack_all <- raster::stack("./Results/raster_stack_all_update.tif")

# convert start to data frame and keep coordinates
# Convert the RasterStack to points
raster_stack_all <- raster::rasterToPoints(raster_stack_all)

# Convert the points to a DataFrame
raster_stack_all_df <- as.data.frame(raster_stack_all, xy = TRUE)


## data clean
# raster_stack_all_df1 <- raster_stack_all_df %>%
#         filter(!is.na(nightlight_trend_global) & !is.na(NDVI_trend_global) & !is.na(PM2.5_trend_global) & co2_trend_global != 0 )

raster_stack_all_df <- raster_stack_all_df %>%
        drop_na()


# giving scores to each indicator
raster_stack_all_df$co2_score <- ifelse(raster_stack_all_df$co2_trend_global <= 0, 1, 0)

raster_stack_all_df$PM2_5_score <- ifelse(raster_stack_all_df$PM2_5_trend_global_update <= 0, 1, 0)

raster_stack_all_df$nightlight_score <- ifelse(raster_stack_all_df$nightlight_trend_global_update > 0, 1, 0)

raster_stack_all_df$NDVI_score <- ifelse(raster_stack_all_df$NDVI_trend_global_update > 0, 1, 0)

raster_stack_all_df$score <- raster_stack_all_df$co2_score + raster_stack_all_df$nightlight_score + raster_stack_all_df$NDVI_score + raster_stack_all_df$PM2_5_score

raster_stack_all_df$score <- as.factor(raster_stack_all_df$score)

saveRDS(raster_stack_all_df, file = "./Results/raster_stack_all_df_update.rds")

# save the score to a raster
raster_score <- raster_stack_all_df %>%
        dplyr::select(x, y, score) %>%
        mutate(score = as.factor(score)) %>%
        raster::rasterFromXYZ(xyz = ., crs = "+proj=longlat +datum=WGS84 +no_defs")


writeRaster(
        x = raster_score,
        filename = "./Results/raster_score_update.tif",
        format = "GTiff",
        overwrite = TRUE
)

```

## 4. plot the raster score

```{r}
p_score <- ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = score)) +
        geom_raster() +
        coord_fixed()+
        scale_fill_manual(values = c( "#e12729",  "#f37324", "#f8cc1b",  "#72b043", "#007f4e"),
                          name="Classification",
                          labels=c("Very Bad","Bad","Medium", "Good", "Very Good" ))+
  theme_minimal() +
  labs(fill="score", x=NULL, y=NULL)

# p_score <- p_score + ggtitle("Classification for PM2.5, CO2, Nightlight and NDVI",
#                   subtitle = "Changes per decade were calculated from 2000 to 2019. A preventive trend will be assigned a positive score of 1, \nwhile a harmful trend will receive a score of 0")

# save the p_score to tif
tiff("./Results/result_plots/plot_score_update.tif", width = 2500, height = 1000, units = "px", res = 300)

# Plot the ggplot
print(p_score)

# Close the TIFF device
dev.off()
```

```{r}
# plot other indictors
# co2
p_co2 <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = as.factor(co2_score))) +
        geom_raster() +
        coord_fixed()+
        scale_fill_manual(values = c( "#e12729",  "#72b043"),
                          name="Trend",
                          labels=c("Increase","Decline"))+
  theme_minimal() +
  labs(fill="co2_score", x=NULL, y=NULL)+
        ggtitle("CO2")
# pm2.5
p_PM25 <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = as.factor(PM2_5_score))) +
        geom_raster() +
        coord_fixed()+
        scale_fill_manual(values = c( "#e12729",  "#72b043"),
                          name="Trend",
                          labels=c("Increase","Decline"))+
  theme_minimal() +
  labs(fill="PM2_5_score", x=NULL, y=NULL)+
        ggtitle("PM2.5")

# nightlight
p_nightlight <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = as.factor(nightlight_score))) +
        geom_raster() +
        coord_fixed()+
        scale_fill_manual(values = c( "#e12729",  "#72b043"),
                          name="Trend",
                          labels=c("Decline","Increase"))+
  theme_minimal() +
  labs(fill="nightlight_score", x=NULL, y=NULL)+
        ggtitle("Nightlight")

# NDVI
p_NDVI <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = as.factor(NDVI_score))) +
        geom_raster() +
        coord_fixed()+
        scale_fill_manual(values = c( "#e12729","#72b043"),
                          name="Trend",
                          labels=c("Decline","Increase"))+
  theme_minimal() +
  labs(fill="NDVI_score", x=NULL, y=NULL)+
        ggtitle("NDVI")

```

```{r}
# plot 4 indictor in one plot
p_all <- cowplot::plot_grid(p_co2, p_PM25, p_nightlight, p_NDVI, ncol = 2, nrow = 2)

# save the p_score to tif
tiff("./Results/result_plots/plot_4indictors_update.tif", width = 5000, height = 2000, units = "px", res = 300)

# Plot the ggplot
print(p_all)

# Close the TIFF device
dev.off()
```

## 5. plot the trend of each indicator with exact value

```{r}
# co2
p_co2_value <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = co2_trend_global_update)) +
        geom_raster() +
        coord_fixed()+
        scale_fill_viridis_c(option = "D")+
  theme_minimal() +
  labs(fill="CO2 trend", x=NULL, y=NULL)+
        ggtitle("CO2")

# pm2.5
p_PM25_value <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = PM2_5_trend_global_update)) +
        geom_raster() +
        coord_fixed()+
        scale_fill_viridis_c(option = "D")+
  theme_minimal() +
  labs(fill="PM2.5 trend", x=NULL, y=NULL)+
        ggtitle("PM2.5")

# nightlight
p_nightlight_value <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = nightlight_trend_global_update)) +
        geom_raster() +
        coord_fixed()+
        scale_fill_viridis_c(option = "D",direction = -1)+
  theme_minimal() +
  labs(fill="Nightlight trend", x=NULL, y=NULL)+
        ggtitle("Nightlight")

# NDVI
p_NDVI_value <-ggplot(data = raster_stack_all_df, aes(x = x, y = y, fill = NDVI_trend_global_update)) +
        geom_raster() +
        coord_fixed()+
        scale_fill_viridis_c(option = "D",direction = -1)+
  theme_minimal() +
  labs(fill="NDVI trend", x=NULL, y=NULL)+
        ggtitle("NDVI")

```

```{r}
# plot 4 indictor in one plot
p_all_value <- cowplot::plot_grid(p_co2_value, p_PM25_value, p_nightlight_value, p_NDVI_value, ncol = 2, nrow = 2)

# save the p_score to tif
tiff("./Results/result_plots/plot_4indictors_value_update.tif", width = 5000, height = 2000, units = "px", res = 300)

# Plot the ggplot
print(p_all_value)

# Close the TIFF device
dev.off()
```


# 6. Category with 16 groups

    1. population with year of 2019
    2. population with year of 2030 (SSP2)
    3. category with 16 groups
    4. calculate the number of population/proportions
    
```{r eval=FALSE, include=FALSE}
# load category data
raster_stack_all_df <- readRDS("./Results/raster_stack_all_df_update.rds")

# create a variable with 16 categories 2^4=16
raster_stack_all_df <- raster_stack_all_df %>%
        mutate(combined_category = paste0(co2_score, PM2_5_score,nightlight_score, NDVI_score)) %>%
        mutate(combined_category = factor(combined_category,levels = c("0000", "1000", "0100", "0010", "0001","1100", "1010","1001","0110","0101","0011", "1110", "1101", "1011", "0111", "1111")))%>%
        mutate(combined_type = as.factor(as.numeric(combined_category)))

# previous coding:
# levels(raster_stack_all_df$combined_category)
#  [1] "0000" "0001" "0010" "0011" "0100" "0101" "0110" "0111" "1000" "1001" "1010" "1011" "1100" "1101" "1110" "1111"

raster_category <- raster_stack_all_df %>%
        dplyr::select(x,y, combined_type) %>%
        raster::rasterFromXYZ(xyz = ., crs = "+proj=longlat +datum=WGS84 +no_defs")

writeRaster(
        x = raster_category,
        filename = "./plots_data/raster_category_16_update.tif",
        format = "GTiff",
        overwrite = TRUE
)

# load data from 2019
year_data_df_2019 <- stack("./plots_data/landscan_global_UN_adjusted_2019.tif")

#year_pop_2019 <- subset(year_data_df_2019, subset = "pop")


### convert it to 0.1 resolution by sum all pop
# Load the raster data
r <- year_data_df_2019

# Define the factor of aggregation
# Current resolution is 0.008333333 (approximately 1/120 degrees), and target is 0.1 degrees.
# Factor is calculated as: 0.1 / 0.008333333 ≈ 12
aggregation_factor <- round(0.1 / res(r)[1])

# Aggregate raster to 0.1-degree resolution by summing values
r_aggregated <- aggregate(r, fact = aggregation_factor, fun = sum, na.rm = TRUE)

year_pop_2019_resample <- resample(r_aggregated, raster_category,  method='ngb')  # using ngb to keep value consistent

cellStats(year_pop_2019_resample,stat = sum)

writeRaster(
        x = year_pop_2019_resample,
        filename = "./plots_data/year_pop_2019_resample_UN_adjusted.tif",
        format = "GTiff",
        overwrite = TRUE
)

# load population 2030

for (i in 3:5) {
        year_pop_ssp_2030 <- raster(paste0("S:\\MNHS-SPHPM-CARE\\GLOBALENVIRONHealth\\Global SSP population data 2010_2100\\1km downscaled projects\\SSP",i,"_1km_netcdf\\SSP",i,"_1km\\ssp",i,"_total_2030.nc4"))

factor <- 0.1 / 0.008333333
year_pop_ssp_2030_0.1 <- aggregate(year_pop_ssp_2030, fact=factor, fun=sum, na.rm=TRUE)


year_pop_ssp_2030_resample <- resample(year_pop_ssp_2030_0.1, raster_category,  method='bilinear')

writeRaster(
        x = year_pop_ssp_2030_resample,
        filename = paste0("./plots_data/year_pop_ssp",i,"_2030_resample.tif"),
        format = "GTiff",
        overwrite = TRUE
)
        
}

```

```{r}
# load population 2019, 2030 and category
pop_2019 <- raster::raster(x = "./plots_data/year_pop_2019_resample_UN_adjusted.tif")

pop_2030_ssp1 <- raster::raster(x = "./plots_data/year_pop_ssp1_2030_resample.tif")
pop_2030_ssp2 <- raster::raster(x = "./plots_data/year_pop_ssp2_2030_resample.tif")
pop_2030_ssp3 <- raster::raster(x = "./plots_data/year_pop_ssp3_2030_resample.tif")
pop_2030_ssp4 <- raster::raster(x = "./plots_data/year_pop_ssp4_2030_resample.tif")
pop_2030_ssp5 <- raster::raster(x = "./plots_data/year_pop_ssp5_2030_resample.tif")

category_type <- raster::raster(x = "./plots_data/raster_category_16_update.tif")


raster_category_pop_2019_2030 <-
                stack(c(category_type, pop_2019, pop_2030_ssp1,pop_2030_ssp2,pop_2030_ssp3,pop_2030_ssp4,pop_2030_ssp5))

names(raster_category_pop_2019_2030) <- c("category", "pop_2019","pop_2030_ssp1","pop_2030_ssp2","pop_2030_ssp3","pop_2030_ssp4","pop_2030_ssp5")

Fig3_df <- raster_category_pop_2019_2030 %>%
        raster::as.data.frame(xy = TRUE) %>%
        filter(is.na(category)==FALSE)

saveRDS(Fig3_df, "./plots_data/Fig_df_update.rds")

# # save as tif
# raster_category_pop_2019_2030 <- st_as_stars(raster_category_pop_2019_2030)
# 
# write_stars(raster_category_pop_2019_2030, "./plots_data/raster_category_pop_2019_2030_update_20241213.tif")

Fig3 <- Fig3_df %>%
        mutate(category = as.factor(category)) %>%
        group_by(category) %>%
        summarise(sum_pop_2019 = sum(pop_2019, na.rm = TRUE),
                  sum_pop_2030_ssp1 = sum(pop_2030_ssp1, na.rm = TRUE),
                  sum_pop_2030_ssp2 = sum(pop_2030_ssp2, na.rm = TRUE),
                  sum_pop_2030_ssp3 = sum(pop_2030_ssp3, na.rm = TRUE),
                  sum_pop_2030_ssp4 = sum(pop_2030_ssp4, na.rm = TRUE),
                  sum_pop_2030_ssp5 = sum(pop_2030_ssp5, na.rm = TRUE)) %>%
        mutate(proportion_2019 = round((sum_pop_2019/sum(sum_pop_2019,na.rm = TRUE))*100,2),
              proportion_2030_ssp1 = round((sum_pop_2030_ssp1/sum(sum_pop_2030_ssp1,na.rm = TRUE))*100,2),
             proportion_2030_ssp2 = round((sum_pop_2030_ssp2/sum(sum_pop_2030_ssp2,na.rm = TRUE))*100,2),
              proportion_2030_ssp3 = round((sum_pop_2030_ssp3/sum(sum_pop_2030_ssp3,na.rm = TRUE))*100,2),
             proportion_2030_ssp4 = round((sum_pop_2030_ssp4/sum(sum_pop_2030_ssp4,na.rm = TRUE))*100,2),
              proportion_2030_ssp5 = round((sum_pop_2030_ssp5/sum(sum_pop_2030_ssp5,na.rm = TRUE))*100,2)
               )

write.csv(Fig3, file = "./plots_data/Fig_3/Fig3_update_ssp1_5_update_UN_adjusted.csv")
```

# 6 Number(proportion) of the best and worst category

    13 rows: Global, continent, income groups, urban and rural
    
    1. national shape files
    2. continent shape files
    3. nation with different income groups
    4. urban /rural categories in 2019

```{r}
# load population, category_type, urbanization, income group
pop_2019 <- raster::raster(x = "./plots_data/year_pop_2019_resample_UN_adjusted.tif")

category_type <- raster::raster(x = "./plots_data/raster_category_16_update.tif")

urban_2019 <- raster::raster(x = "./plots_data/year_urban_2019_resample.tif")

income_group <- read.csv("./plots_data/World_bank_income_group_2019.csv")

pop_2030 <- raster::raster(x = "./plots_data/year_pop_ssp2_2030_resample.tif")

raster_category_pop_urban <-
                stack(c(category_type, pop_2019,pop_2030, urban_2019)) %>%
                as.data.frame(xy = TRUE) 

```

```{r}

# load region raster data for sub-region categorization
region_raster_df <- readRDS("./plots_data/region_raster_df.rds")

sub_region_category_intersect <- raster_category_pop_urban %>%
        left_join(region_raster_df, by = c("x"="X", "y"="Y"))
```

```{r}
table1_df <- sub_region_category_intersect %>%
        filter(is.na(raster_category_16_update)==FALSE)

# link income group
table1_df <- table1_df %>%
        left_join(income_group,by = c("iso_a3"="ISO3"))

# Table 1 calculation
table1_df <- table1_df %>%
        mutate(raster_category_16 = factor(raster_category_16_update, levels = 1:16),
               raster_category_16_update = NULL,
               SUBREGION2 = factor(SUBREGION2, levels = c("Northern America","Latin America and the Caribbean","Northern Europe","Eastern Europe","Western Europe","Southern Europe","Northern Africa","Sub-Saharan Africa","South-eastern Asia","Western Asia","Central Asia","Southern Asia","Eastern Asia","Australia and New Zealand","Other regions in Oceania")),
               REGION = factor(REGION,levels = c("Americas","Europe","Africa","Asia","Oceania")),
               Income2019 = factor(Income2019, levels = c("H","UM","LM","L")),
               urban_2019 = if_else(year_urban_2019_resample == 30, "Urban centre", if_else(year_urban_2019_resample == 10,"Water", "Rural")),
               urban_2019 = factor(urban_2019, levels = c("Urban centre","Rural","Water")))


#saveRDS(table1_df, file = "./plots_data/Table1/table1_df_update.rds")
```


```{r}
# Assuming your dataframe is named df
df <- table1_df %>%
        drop_na()

# Create a summary table with total population for each raster category, their proportions, and total sum
summary_table <- df %>%
        # Add a constant column for global aggregation
        mutate(Global = "Global") %>%
        # Gathering data to long format for easier summarization
        pivot_longer(cols = c("Global", "REGION", "Income2019", "urban_2019"), names_to = "Category", values_to = "Value") %>%
        # Grouping by the new long format categories and raster category
        group_by(Category, Value, raster_category_16) %>%
        # Summarizing with sum of year_pop_2019_resample
        summarise(Total_Population = sum(year_pop_2019_resample,na.rm = TRUE), .groups = 'drop') %>%
        # Calculate total sum across all categories within each group
        group_by(Category, Value) %>%
        mutate(Total_Sum = sum(Total_Population,na.rm = TRUE)) %>%
        # Calculate proportion for each category
        mutate(Proportion = round(Total_Population / Total_Sum*100,2)) %>%
        ungroup() %>%
        # Pivot wider to get separate columns for each raster category
        pivot_wider(names_from = raster_category_16, values_from = c(Total_Population, Proportion), names_sep = "_")

write.csv(summary_table, file = "./plots_data/Table1/summary_table1_update.csv")

```


# 7 rank with nations

        A.Top 15(or 20) countries by number of best category
        B.Top 15(or 20) countries by number of worst category
        C.Top 15(or 20) countries by proportion of best category
        D.Top 15(or 20) countries by proportion of worst category
        
    1. calculate national population (best and worst)
    2. calculate national population proportion (best and worst)
    
```{r}
table1_df <- readRDS("./plots_data/table1_df_updated.rds")

# group by country

Fig4_data <- table1_df %>%
        group_by(iso_a3, Country_name, raster_category_16) %>%
        summarise(sum_pop = sum(year_pop_2019_resample,na.rm = TRUE)) %>%
        group_by(iso_a3,Country_name) %>%
        mutate(proportion = sum_pop/sum(sum_pop,na.rm = TRUE),
               proportion = round(proportion*100,2))

#saveRDS(Fig4_data, file = "./plots_data/Fig4/Fig4_data_update.rds")

# A.Top 15(or 20) countries by number of best category

top15_best_numb <- Fig4_data %>%
        filter(raster_category_16 == 16) %>%
        arrange(desc(sum_pop))
top15_best_numb <- top15_best_numb[1:15,]

# write.csv(top15_best_numb, file = "./plots_data/Fig4/top15_best_numb_update.csv")

# B.Top 15(or 20) countries by number of worst category
top15_worst_numb <- Fig4_data %>%
        filter(raster_category_16 == 1) %>%
        arrange(desc(sum_pop)) 
top15_worst_numb <- top15_worst_numb[1:15,]

# write.csv(top15_worst_numb, file = "./plots_data/Fig4/top15_worst_numb_update.csv")

# C.Top 15(or 20) countries by proportion of best category
top15_best_proportion <- Fig4_data %>%
        filter(raster_category_16 == 16) %>%
        arrange(desc(proportion)) 
top15_best_proportion <- top15_best_proportion[1:15,]

# write.csv(top15_best_proportion, file = "./plots_data/Fig4/top15_best_proportion_update.csv")

# D.Top 15(or 20) countries by proportion of worst category
top15_worst_proportion <- Fig4_data %>%
        filter(raster_category_16 == 1) %>%
        arrange(desc(proportion)) 
top15_worst_proportion <- top15_worst_proportion[1:15,]
# write.csv(top15_worst_proportion, file = "./plots_data/Fig4/top15_worst_proportion_update.csv")

```